---
title: "Triangulation report"
output:
 html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Causal inference and interpretation report

It is often difficult to draw conclusions about causal relationships from observational data. We expect each individual analysis to be biased in some way, so we cannot be confident that the effect estimate generated by the analysis reflects a true causal effect.

However, there are some methods that can be applied to strengthen causal inference from observational data. Although all of these methods will be biased to some extent, some of them have different and unrelated sources of bias. Therefore, if we examine the same causal question by applying multiple methods and then comparing (or *triangulating*) the results, we can be more confident that a causal relationship does (or does not) exist.

```{r, echo=FALSE, results='asis'}
triang_sum <- global_data$triangulation_summaries$triangulation_pregnancy

expclass <- tolower(input$exposure_class_tri)
parent <- tolower(input$exposure_person_tri)
outc <- tolower(input$outcome_subclass2_tri)

triang_sum <- droplevels(triang_sum[(tolower(triang_sum$exposure)==expclass)&
                                            (tolower(triang_sum$outcome_subclass2)==outc)&
                                            (tolower(triang_sum$parent)==parent),])
triang_res <- as.character(min(triang_sum$value[grepl("of evidence",triang_sum$value)]))

```

### Evidence included in this report

This report provides evidence from the following types of analysis (depending on data availability):

-   Multivariable regression of the exposure any time in pregnancy (M)

-   Dose-response (D)

-   Maternal vs partner comparisons (N)

-   Temporal specificity (timing of exposure) (P)

-   Genetic risk scores (G)


This report also gives a little more information to provide further help in interpreting relationships. This further information will not necessarily tell us anything more about causality, but it can help us understand and contextualise the findings.

-   Different exposure definitions 

-   Comparison to socioeconomic position

## The relationship of interest

This report considers:

-   **Exposure class:** `r expclass` during pregnancy
-   **Exposed parent:** `r parent`
-   **Child's outcome:** `r outc`

**Underlying causal hypothesis:** The underlying causal hypothesis is that the `r parent`'s exposure (`r expclass`) during pregnancy causally affects the child's outcome (`r outc`).

**Target trial:** We often rely on observational data because it would be impractical or unethical to conduct an experiment (trial). However, it can be useful to define the design of a hypothetical trial. If we were studying this hypothesis using a randomised trial design, we would recruit participants who do not engage in each behaviour pre-pregnancy (i.e. unexposed pre-pregnancy) and randomise some `r parent`s to be 'exposed' (e.g. smoke, consume alcohol, or consume caffeine) for the entire pregnancy and some `r parent`s to remain unexposed (i.e. to not engage in those health behaviours at all during pregnancy). Then we would compare the outcome of interest in the children of the exposed and unexposed `r parent`s.

## Triangulation summary

**Overall summary** The EPoCH study identified `r triang_res` to support the underlying causal hypothesis.

